Peer review assignment 1 - Reproducible Research (Coursera)
-----------------------------------------------------------

Libraries that are needed are as follows:
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
```

## Opening and tidying the data.
Open activity.zip and read the csv, explore the data.
```{r}
csv <- unzip('activity.zip')
df <- read.csv(csv)
head(df)
dim(df)
str(df)
```

Put the date variable in a date format, and time in a POSIX format.
```{r}
df$date <- as.Date(df$date)
df$interval <- as.POSIXct(sprintf("%04d",df$interval), format="%H%M") 
str(df)
```

## Number of steps per day.
We calculate the total number of steps per day.
```{r}
df_day <- df %>% group_by(date) %>% 
    summarise(sum(steps, na.rm = TRUE), .groups='drop')
df_day$steps <- df_day$`sum(steps, na.rm = TRUE)`
```

Plot the total number of steps per day with a ggplot histogram.
```{r}
ggplot(df_day) +
    geom_bar(aes(date, steps), stat = 'identity', position='identity', fill = 'royalblue')
```

Calculate the mean and median steps per day:
```{r}
mean <- round(mean(df_day$steps, na.rm = TRUE), digits = 0)
median <- round(median(df_day$steps, na.rm = TRUE), digits = 0)
```
The mean number of steps per day is `r mean[1]` and the median number of steps per day is `r median[1]`.

## Number of steps per time interval.
First, we calculate the average number of steps per time interval.
```{r}
df_time <- df %>% group_by(interval) %>% 
    summarise(mean(steps, na.rm = TRUE), .groups='drop')
df_time$steps <- df_time$`mean(steps, na.rm = TRUE)`
head(df_time)
```

Time series plot for the average number of steps per time interval:
```{r}
ggplot(df_time) +
    geom_line(aes(interval, steps), color = 'royalblue') +
    scale_x_datetime(date_label = "%H", date_breaks = "1 hour") +
    labs(title='Average number of steps per time interval')
```

What is the time interval with the most average steps?
```{r}
begin <- format(df_time$interval[which(df_time$steps==max(df_time$steps, na.rm = TRUE))], format = "%H:%M")
end <- format(df_time$interval[which(df_time$steps==max(df_time$steps, na.rm = TRUE))]+(5*60), format = "%H:%M")
```
The time interval between `r begin[1]` and `r end[1]` has the most average steps.

## Imputing the missing values
First, we calculate the number of missing values in this data set.
```{r}
missings <- sum(is.na.data.frame(df))
```
There are `r missings[1]` missing values in the data set.

Now we explore where the values are missing - per time interval and per date.
```{r}
df_missing <- df[is.na(df$steps),]
ggplot(df_missing) +
    geom_bar(aes(interval)) +
    scale_x_datetime(date_label = "%H", date_breaks = "1 hour")
ggplot(df_missing) +
    geom_bar(aes(date))
```

The missing values originate from 8 dates that miss all values.

Imputing missing values with mean for the associated time interval:
```{r}
df_imp <- df
for (a in which(is.na(df_imp$steps))) {
    int <- format(df_imp$interval[a], format="%Y-%m-%d %H:%M:%S")
    mean_int <- mean(df_imp$steps[which(df_imp$interval == int)], na.rm=TRUE)
    df_imp$steps[a] <- mean_int
}
missings_imp <- sum(is.na.data.frame(df_imp)) # Checking if there are missings left
print(ifelse(missings_imp==0, 'There are no missing values left', 'There are missing values left'))
```
Now we can calculate the new total steps per day, and plot the number of steps.
```{r}
df_day2 <- df_imp %>% group_by(date) %>% 
    summarise(sum(steps, na.rm = TRUE), .groups='drop')
df_day2$steps <- df_day2$`sum(steps, na.rm = TRUE)`

ggplot(df_day2) +
    geom_bar(aes(date, steps), stat = 'identity', position='identity', fill = 'royalblue')
```

Calculating the new mean and median for the imputed data set:
```{r}
mean_imp <- round(mean(df_day2$steps, na.rm = TRUE), digits = 0)
median_imp <- round(median(df_day2$steps, na.rm = TRUE), digits = 0)
```
The mean was `r mean[1]` and became `r mean_imp [1]` after imputing; the median was `r median [1]` and became `r median_imp[1]` after imputing.

## Differences between week and weekend days
To examine the differences between weekdays and weekend days, we make a new factor variable 'week'.
```{r}
df_imp$week <- ifelse(weekdays(df_imp$date)=='zaterdag' |weekdays(df_imp$date)=='zondag', 'weekend', 'weekday')
df_imp$week <- as.factor(df_imp$week)
summary(df_imp$week)
```

And calculate the averages over time intervals for weekdays and weekend days separately.
```{r}
df_time2 <- df_imp %>% group_by(interval, week) %>% 
    summarise(mean(steps, na.rm = TRUE), .groups='drop')
df_time2$steps <- df_time2$`mean(steps, na.rm = TRUE)`
head(df_time2)
```

Finally, we plot the average steps in time series plots for weekdays vs weekend days.
```{r}
ggplot(df_time2) +
    geom_line(aes(interval, steps), color = 'royalblue') +
    scale_x_datetime(date_label = "%H", date_breaks = "4 hours") +
    facet_wrap(~week) +
    labs(title='Average number of steps per time interval - weekdays versus weekend days')
```
